---
title: "Case1- Cyclistic"
author: "Shahar Mazur"
date: "2022-08-24"
output: html_document
---
# To manipulate date functions
# To create final visualization
# To calculate distances on the map
# To scale th ggplot

knitr::opts_chunk$set(echo = TRUE)



install.packages("lubridate") 
install.packages("ggplot2") 
install.packages("geosphere") 
install.packages("scales") 
install.packages("ggmap")


```{r setup, include=FALSE}
library(geosphere)
library(lubridate)
library(ggplot2)
library(tidyverse)
library(scales)
library(ggmap)
```

# As Im working on my desktop, making sure the working dir.
## Setting my data dir 
```{r}
getwd() 
setwd("/Users/mazor/Documents/Case1-Cyclistic/Data") 
list.files(path = "../Data")
```
# Reading in the entire Dataset - 1 Year's worth of traveling data
```{r}
tripdata_202107 <- read.csv("../data/202107-divvy-tripdata.csv")
tripdata_202108 <- read.csv("../data/202108-divvy-tripdata.csv")
tripdata_202109 <- read.csv("../data/202109-divvy-tripdata.csv")
tripdata_202110 <- read.csv("../data/202110-divvy-tripdata.csv")
tripdata_202111 <- read.csv("../data/202111-divvy-tripdata.csv")
tripdata_202112 <- read.csv("../data/202112-divvy-tripdata.csv")
tripdata_202201 <- read.csv("../data/202201-divvy-tripdata.csv")
tripdata_202202 <- read.csv("../data/202202-divvy-tripdata.csv")
tripdata_202203 <- read.csv("../data/202203-divvy-tripdata.csv")
tripdata_202204 <- read.csv("../data/202204-divvy-tripdata.csv")
tripdata_202205 <- read.csv("../data/202205-divvy-tripdata.csv")
tripdata_202206 <- read.csv("../data/202206-divvy-tripdata.csv")
```


#.................#
#Checking My Data #
#.................#
# Checking coulmn names are identical before joining all data
```{r}
colnames(tripdata_202107)
colnames(tripdata_202108)
colnames(tripdata_202109)
colnames(tripdata_202110)
colnames(tripdata_202111)
colnames(tripdata_202112)   
colnames(tripdata_202201)
colnames(tripdata_202202)
colnames(tripdata_202203)
colnames(tripdata_202204)
colnames(tripdata_202205)
colnames(tripdata_202206)
```




# Joining all data to one dataset to make future processing more simple
```{r}
Total_Trip_Data <- bind_rows(tripdata_202107
                             ,tripdata_202108
                             ,tripdata_202109
                             ,tripdata_202110
                             ,tripdata_202111
                             ,tripdata_202112
                             ,tripdata_202201
                             ,tripdata_202202
                             ,tripdata_202203
                             ,tripdata_202204
                             ,tripdata_202205
                             ,tripdata_202206)
```


```{r}
print(unique(Total_Trip_Data$rideable_type)) 
```
### Consistent with 3 types of bikes - "docked_bike"   "classic_bike"  "electric_bike" - No Nulls
```{r}
print(unique(Total_Trip_Data$member_casual))          
```
### Consistent with 2 types of users - "casual" "member" - No Nulls
```{r include=FALSE}
print(unique(Total_Trip_Data$start_station_name))                                         
print(unique(Total_Trip_Data$end_station_name))
```
### There are 22 more ending stations than starting stations
```{r include=FALSE}
print(unique(Total_Trip_Data$start_station_id))
print(unique(Total_Trip_Data$end_station_id))
```
### There are 14 more ending stations ID's than starting ID

### Data contain a lot of un-needed data, as charging stations, repair stations, mobile stations .
```{r}
print(mean(Total_Trip_Data$start_lat))
print(mean(Total_Trip_Data$start_lng))
print(mean(Total_Trip_Data$end_lat))     
print(mean(Total_Trip_Data$end_lng))     
```

##### mean(Total_Trip_Data$end_lng came out as NA
##### mean(Total_Trip_Data$end_lat came out as NA
##### A quick check to see if there are NULLs in a numeric data frames
##### It shows some of the bikes never arrived to an end station (theft, broke) either way the data will interfere with our results and should be removed.


##### 774 bikes has not arrived to an end station by members.
##### 3545 bikes has not arrived to an end station by casual users.
##### Total of 4319 bikes has not arrived to an end station.




# Organizing My Data


```{r}
Total_Trip_Data <- na.omit(Total_Trip_Data) 
```
##### removes all null cells that might interfere with calculations. - Removes 5,374 rows

```{r include=FALSE}
Total_Trip_Data <- rename(Total_Trip_Data
                  ,start_time = "started_at"
                  ,end_time = "ended_at")
```

### extract year, month, date, day of week .

```{r}
Total_Trip_Data <- Total_Trip_Data %>%
  mutate(year = format(as.Date(start_time), "%Y")) %>% 
  mutate(month = format(as.Date(start_time), "%B")) %>% 
  mutate(date = format(as.Date(start_time), "%d")) %>% 
  mutate(day_of_week = format(as.Date(start_time), "%A")) %>% 
  mutate(ride_length = difftime(end_time, start_time))  

Total_Trip_Data <- Total_Trip_Data %>%   
  mutate(ride_length = as.numeric(ride_length))

```
### Creating back up
```{r eval=FALSE, include=FALSE}
Total_Trip_Data -> Total_Trip_Data_temp ## creating back up #1
```

# Adding a column to calculate each trip duration
```{r}
Total_Trip_Data <- mutate(Total_Trip_Data, tripduration = int_length(interval(ymd_hms(Total_Trip_Data$start_time), ymd_hms(Total_Trip_Data$end_time))))
```

# Calculating aerial distance of the trip.
```{r}
Total_Trip_Data$Ariel_distance <- distGeo(matrix(c(Total_Trip_Data$start_lng, Total_Trip_Data$start_lat), ncol = 2), matrix(c(Total_Trip_Data$end_lng, Total_Trip_Data$end_lat), ncol = 2))


Total_Trip_Data <- Total_Trip_Data[!(Total_Trip_Data$Ariel_distance <= 0),]
```
# Cleaning rides that were picked up and returned to the same station. 
#### Cleared 366,276 rows (this is a data collection issue)

```{r}
Total_Trip_Data <- Total_Trip_Data[!(Total_Trip_Data$Ariel_distance >=30000),] 
```
## Cleaning  unusual rides, over 30KM  (cleared 10 rows)

```{r}
Total_Trip_Data <- Total_Trip_Data[!(Total_Trip_Data$tripduration <= 0),] 
```
## Cleaning rows with issue as trip duration =<0 , cleared 484 rows


# Creating Backup number 2
```{r eval=FALSE, include=FALSE}
Total_Trip_Data_temp2 <- Total_Trip_Data ## creating back up #2
```


## Analyzing
##### This analize is going to be without assuming casual riders go to touristic spots, in the 2nd analysis wi will remove all casual riders ending stations that are located in a city touristic area so we can consider only local casual riders.

```{r}
Total_Trip_Data %>% 
  group_by(rideable_type) %>% 
  summarise(ride_count = length(ride_id), ride_percentage = (length(ride_id) / nrow(Total_Trip_Data)) * 100)
```
# what is "docked_bike" ?
### Contacted Divvy , the answer was that Docked_bike are equal to Classic_bike.
```{r}
Total_Trip_Data['rideable_type'][Total_Trip_Data['rideable_type'] =="docked_bike"] <- "classic_bike"
print(unique(Total_Trip_Data$rideable_type))
```
### Corrected to 2 types of bikes -  "classic_bike"  "electric_bike" - No Nulls


```{r}
Total_Trip_Data %>% 
  group_by(member_casual) %>% 
  summarise(ride_count = length(ride_id), ride_percentage = (length(ride_id) / nrow(Total_Trip_Data)) * 100)
```


```{r}
Total_Trip_Data %>% 
  group_by(member_casual) %>% 
  summarise(longest_ride_MINUTES = max(tripduration/60),Average_ride_MINUTES = mean(tripduration/60),Shortest_ride_MINUTES = min(tripduration/60))
```



```{r}  
Total_Trip_Data %>%
ggplot(mapping = aes(x = member_casual, fill=member_casual)) +
  geom_bar() +
  labs( x="Casuals vs Members", y="Number Of Rides", title= "Distrebution between Members & Casual users - Tourists included.",alt = "User Type", caption = "**This dataset contain tourists**") +
  scale_y_continuous(labels = comma) +
  scale_fill_manual("legend", values = c("member" = "#62B2E0", "casual" = "#E06287")) +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_text(aes(label = scales::percent((..count..)/sum(..count..))), stat = "count", vjust = 12, colour = "Black")
```
## 42.4%
## 57.6%



## Sorting "day_of_week" by the days of the week 
```{r}
Total_Trip_Data$day_of_week <- factor(Total_Trip_Data$day_of_week, 
                               levels = as.character(wday(c(2:7,1), label=TRUE, abbr=FALSE)))
```

## Sorting "month" by the months of the year
```{r}
Total_Trip_Data$month <- factor(Total_Trip_Data$month, 
                                      levels = as.character(month(c(7:12,1:6), label=TRUE, abbr=FALSE)))
```
```{r}
ggplot(Total_Trip_Data, aes(x=day_of_week,fill = member_casual)) +
  geom_bar(position = "dodge") +
  scale_y_continuous(labels = comma)+
  theme(plot.title = element_text(hjust = 0.5)) +
  labs( x="Day Of The Week", y="Number Of Rides", title= "Members & Casual users use by day of the week - With Tourists",alt = "User Type", caption = "**This dataset contain tourists**") +
  scale_fill_manual("legend", values = c("member" = "#62B2E0", "casual" = "#E06287"))
```
```{r}
ggplot(Total_Trip_Data, aes(x=month,fill = member_casual)) +
  geom_bar(position = "dodge") +
  scale_y_continuous(labels = comma)+
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual("legend", values = c("member" = "#62B2E0", "casual" = "#E06287")) +
  labs(x="July 2021 to June 2022", y="Number Of Rides", title= "Members & Casual users use by month.",caption = "**This dataset contain tourists**")
```


### Analyzing these 2 tablets shows us there are too many tourists in the dataset, and the should not be considered in the research.

# Creating a backup for the cleaning process
```{r}
Total_Trip_Data -> Total_Trip_Data_temp2
write.csv(Total_Trip_Data,"C:\\Users\\mazor\\Documents\\Case1-Cyclistic\\Data\\Final_Dataset_with_tourists.csv", row.names = FALSE)
```

# Removing stations that are considered by the company as " for tourism"
```{r}
Total_Trip_Data_no_tourists <- Total_Trip_Data[!(Total_Trip_Data$end_station_name=="Streeter Dr & Grand Ave" 
                                                  | Total_Trip_Data$end_station_name=="Ft Dearborn Dr & 31st St"
                                                  | Total_Trip_Data$end_station_name=="South Shore Dr & 71st St"
                                                  | Total_Trip_Data$end_station_name=="Museum of Science and Industry"
                                                  | Total_Trip_Data$end_station_name=="Shore Dr & 55th St"
                                                  | Total_Trip_Data$end_station_name=="Fort Dearborn Dr & 31st St"
                                                  | Total_Trip_Data$end_station_name=="Burnham Harbor"
                                                  | Total_Trip_Data$end_station_name=="Shedd Aquarium"
                                                  | Total_Trip_Data$end_station_name=="Field Museum"
                                                  | Total_Trip_Data$end_station_name=="Adler Planetarium"
                                                  | Total_Trip_Data$end_station_name=="Lake Shore Dr & Monroe St"
                                                  | Total_Trip_Data$end_station_name=="DuSable Harbor"
                                                  | Total_Trip_Data$end_station_name=="Lake Park Ave & 53rd St"
                                                 | Total_Trip_Data$end_station_name=="Lake Shore Dr & North Blvd"
                                                 | Total_Trip_Data$end_station_name=="Clark St & Armitage Ave"
                                                 | Total_Trip_Data$end_station_name=="Cannon Dr & Fullerton Ave"
                                                 | Total_Trip_Data$end_station_name=="Lake Shore Dr & Belmont Ave"
                                                 | Total_Trip_Data$end_station_name=="Montrose Harbor"
                                                 | Total_Trip_Data$end_station_name=="Michigan Ave & Oak St"
                                                 | Total_Trip_Data$end_station_name=="Marshfield Ave & Courtland St"
                                                 | Total_Trip_Data$end_station_name=="Damen Ave & Courtland St"
                                                 | Total_Trip_Data$end_station_name=="Milwaukee Ave & Wabansia St"
                                                 | Total_Trip_Data$end_station_name=="Western Ave & Winnebago Ave"
                                                 | Total_Trip_Data$end_station_name=="Albany Ave & Bloomingdale Ave"
                                                 | Total_Trip_Data$end_station_name=="Theater on the Lake"
                                                 | Total_Trip_Data$start_station_name=="Streeter Dr & Grand Ave" 
                                                 | Total_Trip_Data$start_station_name=="Ft Dearborn Dr & 31st St"
                                                 | Total_Trip_Data$start_station_name=="South Shore Dr & 71st St"
                                                 | Total_Trip_Data$start_station_name=="Museum of Science and Industry"
                                                 | Total_Trip_Data$start_station_name=="Shore Dr & 55th St"
                                                 | Total_Trip_Data$start_station_name=="Fort Dearborn Dr & 31st St"
                                                 | Total_Trip_Data$start_station_name=="Burnham Harbor"
                                                 | Total_Trip_Data$start_station_name=="Shedd Aquarium"
                                                 | Total_Trip_Data$start_station_name=="Field Museum"
                                                 | Total_Trip_Data$start_station_name=="Adler Planetarium"
                                                 | Total_Trip_Data$start_station_name=="Lake Shore Dr & Monroe St"
                                                 | Total_Trip_Data$start_station_name=="DuSable Harbor"
                                                 | Total_Trip_Data$start_station_name=="Lake Park Ave & 53rd St"
                                                 | Total_Trip_Data$start_station_name=="Lake Shore Dr & North Blvd"
                                                 | Total_Trip_Data$start_station_name=="Clark St & Armitage Ave"
                                                 | Total_Trip_Data$start_station_name=="Cannon Dr & Fullerton Ave"
                                                 | Total_Trip_Data$start_station_name=="Lake Shore Dr & Belmont Ave"
                                                 | Total_Trip_Data$start_station_name=="Montrose Harbor"
                                                 | Total_Trip_Data$start_station_name=="Michigan Ave & Oak St"
                                                 | Total_Trip_Data$start_station_name=="Marshfield Ave & Courtland St"
                                                 | Total_Trip_Data$start_station_name=="Damen Ave & Courtland St"
                                                 | Total_Trip_Data$start_station_name=="Milwaukee Ave & Wabansia St"
                                                 | Total_Trip_Data$start_station_name=="Western Ave & Winnebago Ave"
                                                 | Total_Trip_Data$start_station_name=="Albany Ave & Bloomingdale Ave"
                                                 | Total_Trip_Data$start_station_name=="Theater on the Lake"),]
```
### Removing these stations that are considered as "touristic" removed 569,940 rows, can assume there are more tourists that that a year. but personal information is needed to make sure.
```{r}
Total_Trip_Data_no_tourists <- Total_Trip_Data_no_tourists[!(Total_Trip_Data_no_tourists$Ariel_distance >=30000),] 
Total_Trip_Data_no_tourists <- Total_Trip_Data_no_tourists[!(Total_Trip_Data_no_tourists$tripduration <= 0),]
```


#### This is the last version and change to the dataframe, I work on a laptop so to clear out memory space I will export my dataset into a CSV file, clear the workframe are from old datasets and re-load only what I need for graphic use.

# Backup and reload section
write.csv(Total_Trip_Data_no_tourists,"C:\\Users\\mazor\\Documents\\Case1-Cyclistic\\Data\\Final_Dataset_Without_Tourists_bystation.csv", row.names = FALSE)
Total_Trip_Data_no_tourists <- read.csv("../Data/Final_Dataset_Without_Tourists_bystation.csv") ## ## Don't forget to load libraries.

```{r}
Total_Trip_Data_no_tourists %>% 
  group_by(member_casual) %>% 
  summarise(ride_count = length(ride_id), ride_percentage = (length(ride_id) / nrow(Total_Trip_Data_no_tourists)) * 100)
```

# Significant change from earlier results 
```{r}
Total_Trip_Data_no_tourists %>% 
  group_by(member_casual) %>% 
  summarise(longest_ride_MINUTES = max(tripduration/60),Average_ride_MINUTES = mean(tripduration/60),Shortest_ride_MINUTES = min(tripduration/60))
```


# Starting to plot the data
```{r}
Total_Trip_Data_no_tourists %>%
  ggplot(mapping = aes(x = member_casual, fill=member_casual)) +
  geom_bar() +
  labs( x="Casuals vs Members", y="Number Of Rides", title= "Distrebution between Members & Casual users - Without tourists.",alt = "User Type", caption = "**This dataset contain NO tourists**") +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_y_continuous(labels = comma) +
  scale_fill_manual("legend", values = c("member" = "#62B2E0", "casual" = "#E06287"))+
  geom_text(aes(label = scales::percent((..count..)/sum(..count..))), stat = "count", vjust = 12, colour = "Black")
```


## 40.1%
## 59.9%

## Sorting "day_of_week" by the days of the week 
```{r}
Total_Trip_Data_no_tourists$day_of_week <- factor(Total_Trip_Data_no_tourists$day_of_week, 
                                      levels = as.character(wday(c(2:7,1), label=TRUE, abbr=FALSE)))
```

## Sorting "month" by the months of the year
```{r}
Total_Trip_Data_no_tourists$month <- factor(Total_Trip_Data_no_tourists$month, 
                                levels = as.character(month(c(7:12,1:6), label=TRUE, abbr=FALSE)))
```

```{r}
ggplot(Total_Trip_Data_no_tourists, aes(x=rideable_type,fill = member_casual)) +
  geom_bar(position = "dodge") +
  scale_y_continuous(labels = comma)+
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual("legend", values = c("member" = "#62B2E0", "casual" = "#E06287")) +
  labs(x="Bike Type", y="Number Of Rides", title= "Members & Casual users use by type of bike.",caption = "**This dataset contain NO tourists**")

ggplot(Total_Trip_Data_no_tourists, aes(x=day_of_week,fill = member_casual)) +
  geom_bar(position = "dodge") +
  scale_y_continuous(labels = comma)+
  theme(plot.title = element_text(hjust = 0.5)) +
  labs( x="Day Of The Week", y="Number Of Rides", title= "Members & Casual users use by day of the week - Without Tourists",alt = "User Type", caption = "**This dataset contain NO tourists**") +
  scale_fill_manual("legend", values = c("member" = "#62B2E0", "casual" = "#E06287"))



ggplot(Total_Trip_Data_no_tourists, aes(x=month,fill = member_casual)) +
  geom_bar(position = "dodge") +
  scale_y_continuous(labels = comma)+
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual("legend", values = c("member" = "#62B2E0", "casual" = "#E06287")) +
  labs(x="July 2021 to June 2022", y="Number Of Rides", title= "Members & Casual users use by month.",caption = "**This dataset contain NO tourists**")
```

backup point
Total_Trip_Data_temp3 <- Total_Trip_Data_no_tourists ## back up





```{r}
Total_Trip_Data_no_tourists$hour <-format(as.POSIXct(Total_Trip_Data_no_tourists$start_time), format = "%H")


Total_Trip_Data_no_tourists %>%
  ggplot(aes(x=hour, fill=member_casual)) +
  geom_bar(position = "dodge") +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x="Hour of the day", title="Cyclistic's bike demand per hour by day of the week") +
  scale_fill_manual("legend", values = c("member" = "#62B2E0", "casual" = "#E06287")) +
  facet_wrap(~ day_of_week)


Examin_weekdays <- Total_Trip_Data_no_tourists[!((Total_Trip_Data_no_tourists$day_of_week =="Sunday") |(Total_Trip_Data_no_tourists$day_of_week =="Saturday")),]
```
## Needed to show weekend vs weekdays example
```{r}
Examin_weekdays %>%
  ggplot(aes(x=hour, fill=member_casual)) +
  geom_bar(position = "dodge") +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x="Hour of the day", title="Cyclistic's bike demand per hour on the weekdays") +
  scale_fill_manual("legend", values = c("member" = "#62B2E0", "casual" = "#E06287"))+
  facet_wrap(~ day_of_week,nrow =2)


Examin_weekend <- Total_Trip_Data_no_tourists[(Total_Trip_Data_no_tourists$day_of_week =="Sunday") |(Total_Trip_Data_no_tourists$day_of_week =="Saturday"),]

Examin_weekend %>%
  ggplot(aes(x=hour, fill=member_casual)) +
  geom_bar(position = "dodge") +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x="Hour of the day", title="Cyclistic's bike demand per hour on the weekend") +
  scale_fill_manual("legend", values = c("member" = "#62B2E0", "casual" = "#E06287"))+
  facet_wrap(~ day_of_week)
```
# Exporting files to use in Power BI
```{r}
write.csv(Examin_weekend,"C:\\Users\\mazor\\Documents\\Case1-Cyclistic\\Data\\examine_weekends.csv", row.names = FALSE)
write.csv(Examin_weekdays,"C:\\Users\\mazor\\Documents\\Case1-Cyclistic\\Data\\Examin_weekdays.csv", row.names = FALSE)
```




### Lets check the coordinates data of the rides.
### Adding a new data frame only for the most popular routes >150 rides
### as we are comparing we should first split betwenn members and casual users then look at the 150 most driven routes
```{r}
casual_riders_dt <- Total_Trip_Data_no_tourists %>% filter(member_casual == "casual")
member_riders_dt <- Total_Trip_Data_no_tourists %>% filter(member_casual == "member")
```
### Exporting files to use in Power BI
```{r}
write.csv(casual_riders_dt,"C:\\Users\\mazor\\Documents\\Case1-Cyclistic\\Data\\casual_riders_dt.csv", row.names = FALSE)
write.csv(member_riders_dt,"C:\\Users\\mazor\\Documents\\Case1-Cyclistic\\Data\\member_riders_dt.csv", row.names = FALSE)
```

```{r}
casual_riders <- casual_riders_dt %>% 
  filter(start_lng != end_lng & start_lat != end_lat) %>%
  group_by(start_lng, start_lat, end_lng, end_lat, member_casual, Ariel_distance,rideable_type) %>%
  summarise(total_rides = n(),.groups="drop") %>%
  filter(total_rides > 150)

member_riders <- member_riders_dt %>% 
  filter(start_lng != end_lng & start_lat != end_lat) %>%
  group_by(start_lng, start_lat, end_lng, end_lat, member_casual, Ariel_distance,rideable_type) %>%
  summarise(total_rides = n(),.groups="drop") %>%
  filter(total_rides > 150)
```


# now lets create two different data frames depending on rider type (member_casual)

```{r}
chicago <- c(left = -87.700424, bottom = 41.74, right = -87.554855, top = 42.04)

chicago_map <- get_stamenmap(bbox = chicago, zoom = 12, maptype = "terrain")

total_ride_colors <- colorRampPalette(c("blue","green"))
favorite <- total_ride_colors(200)
```

# Maps on casual riders
```{r}
ggmap(chicago_map,darken = c(0.1, "white")) +
  geom_point(casual_riders, mapping = aes(x = start_lng, y = start_lat, color=total_rides), size = 2) +
  coord_fixed(0.8) +
  scale_colour_gradientn(colors = favorite)+
  labs(title = "Most used routes by Casual riders",x=NULL,y=NULL)
```

# Map on member riders
```{r}
ggmap(chicago_map,darken = c(0.1, "white")) +
  geom_point(member_riders, mapping = aes(x = start_lng, y = start_lat, color=total_rides), size = 2) +  
  coord_fixed(0.8) +
  scale_colour_gradientn(colors = favorite)+
  labs(title = "Most used routes by Member riders",x=NULL,y=NULL)
```



